---
title: "Environmental Performance Index Case Study"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Environmental Performance Index Case Study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Role of the case study

The completed thesis analysed the 2024 Environmental Performance Index corpus
for 180 countries. After missingness screening, low-variance filtering, and
imputation, 70 predictor indicators were retained alongside the EPI response.
The `.old` columns formed the training data and the aligned `.new` columns
formed the test data.

Version 0.3.0 includes the supplied processed predictor and response matrices.
The paired predictor matrices contain the same 70 named indicators and no
missing values. Selected supplied output tables are retained separately as
provenance fixtures.

## Bundled processed inputs

```{r}
read_epi_file <- function(name) {
  utils::read.csv(
    system.file("extdata", "epi", name, package = "risdr"),
    check.names = FALSE
  )
}

X_train <- read_epi_file("X_old.csv")
X_test <- read_epi_file("X_new.csv")
y_train <- read_epi_file("y_old.csv")[[1L]]
y_test <- read_epi_file("y_new.csv")[[1L]]

stopifnot(
  nrow(X_train) == 180L,
  nrow(X_test) == 180L,
  ncol(X_train) == 70L,
  identical(names(X_train), names(X_test)),
  length(y_train) == nrow(X_train),
  length(y_test) == nrow(X_test)
)
```

## Original import and alignment protocol

The transformation from the original EPI corpus should follow the structure
below. It is not evaluated because the single source corpus is separate from
the processed analysis matrices.

```{r raw-data-protocol, eval=FALSE}
library(dplyr)
library(readr)

epi_data <- read_csv("data-raw/epi2024results.csv", show_col_types = FALSE)

epi_old <- epi_data |>
  select(iso, country, ends_with(".old")) |>
  rename_with(\(x) sub("[.]old$", "", x))

epi_new <- epi_data |>
  select(iso, country, ends_with(".new")) |>
  rename_with(\(x) sub("[.]new$", "", x))

X_old_full <- epi_old |> select(-iso, -country)
X_new_full <- epi_new |> select(-iso, -country)

stopifnot(identical(names(X_old_full), names(X_new_full)))

missing_prop_old <- colMeans(is.na(X_old_full))
keep_names <- names(missing_prop_old[missing_prop_old < 0.40])

X_old_reduced <- X_old_full[, keep_names, drop = FALSE]
X_new_reduced <- X_new_full[, keep_names, drop = FALSE]

training_variance <- vapply(X_old_reduced, var, numeric(1), na.rm = TRUE)
keep_names <- names(training_variance[training_variance > 1e-6])

X_old_reduced <- X_old_reduced[, keep_names, drop = FALSE]
X_new_reduced <- X_new_reduced[, keep_names, drop = FALSE]

X_old_imputed <- VIM::kNN(X_old_reduced, k = 5, imp_var = FALSE)
X_new_imputed <- VIM::kNN(X_new_reduced, k = 5, imp_var = FALSE)

y_train <- X_old_imputed$EPI
X_train <- X_old_imputed |> select(-EPI)
y_test <- X_new_imputed$EPI
X_test <- X_new_imputed |> select(-EPI)
```

Filtering and imputation choices must be estimated from the training data and
then applied to the test data. Re-estimating preprocessing independently on the
test data would weaken the interpretation of external prediction.

## Fitting template

```{r fitting-template, eval=FALSE}
fit_epi <- fit_risdr(
  X = X_train,
  y = y_train,
  sdr_method = "phd",
  cov_method = "oas",
  nslices = 6,
  d_max = 10,
  selector = "cicomp",
  standardize = TRUE,
  stabilize = TRUE
)

epi_prediction <- predict(fit_epi, X_test)
evaluate_prediction(y_test, epi_prediction, d = fit_epi$d)
```

The thesis primary predictive comparison reports pHd as its best method, with
approximately RMSE 2.905, MAE 2.321, `R^2` 0.935, adjusted `R^2`
0.933, and correlation 0.967. These values belong to that designated primary
comparison.

## Broader method-covariance audit

The package fixture below records a separate comparison across four SDR methods
and five covariance estimators. It should not be conflated with the primary
thesis comparison because the model grid and selected dimensions differ.

```{r}
comparison_path <- system.file(
  "extdata",
  "epi",
  "epi_sdr_covariance_comparison_tidy.csv",
  package = "risdr"
)
comparison <- utils::read.csv(comparison_path)

comparison[
  order(comparison$RMSE),
  c(
    "sdr_method", "cov_method", "selected_d", "condition_number",
    "RMSE", "MAE", "R2", "Adjusted_R2", "Correlation"
  )
]
```

This audit demonstrates why predictive performance, structural dimension, and
covariance conditioning must be reported together. A method may predict well
under one estimator while retaining a different number of directions under
another.

## Repeated cross-validation

```{r}
cv_path <- system.file(
  "extdata",
  "epi",
  "epi_repeated_cv_summary_tidy.csv",
  package = "risdr"
)
cv_summary <- utils::read.csv(cv_path)

cv_summary[
  order(cv_summary$mean_RMSE),
  c(
    "sdr_method", "cov_method", "n_success", "mean_selected_d",
    "mean_RMSE", "sd_RMSE", "mean_Adjusted_R2", "mean_Correlation"
  )
]
```

The repeated assessment identifies SIR with ridge and SIR with OAS as highly
competitive combinations. This complements, rather than replaces, the fixed
old-to-new comparison.

## Interpretation discipline

The EPI case study supports three distinct conclusions:

1. Prediction, subspace recovery, and structural dimension recovery are
   different objectives.
2. Covariance regularisation can materially improve numerical conditioning.
3. A single overall winner should not be declared without naming the objective,
   validation design, and covariance estimator.

## References
